---
title: "Complex Data lab 3"
author: "Anna Zaleska"
header-includes:
   - \usepackage{xcolor}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme)
library(lattice)
```

```{r}
wtloss <- read.table("../data/weightloss.dat",
header=F)
## Give names to variables
names(wtloss) <- c("id", paste("y", 1:4, sep=""), "program")
## Univariate format
wtloss.uni <- data.frame(id=rep(wtloss$id, each=4),
wgtloss=as.numeric(t(as.matrix(wtloss[,2:5]))),
program=rep(wtloss$program, each=4),
month=seq(0,9,3),
time.cat=rep(1:4))
wtloss.uni$prog.fac <- factor(wtloss.uni$program, labels=c("1:encourage", "2:none"))
attach(wtloss.uni)
wgt.mean <- tapply(wgtloss, list(month, program), mean)
wgt.sd <- tapply(wgtloss, list(month, program), sd)
#wgt.mean
#wgt.sd
detach(wtloss.uni)
```

```{r}

 # Use lattice package
#xyplot(wgtloss~month|prog.fac, type='l',groups=id,data=wtloss.uni)
```

The most general mean models with different covariance classes:

```{r}

## CATEGORICAL time, UN, REML
wtloss.un.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | month),
data=wtloss.uni)
#summary(wtloss.un.cat)

## CATEGORICAL time, CS, REML
wtloss.cs.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni)
#summary(wtloss.cs.cat)

## CATEGORICAL time, AR1, REML
wtloss.ar1.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corAR1(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni)
#summary(wtloss.ar1.cat)
```

### Unstructured vs compound symmetry
$H_0:$ compound symmetry is adequate for the data \newline
$H_1:$ unstructred is required
```{r}
anova(wtloss.un.cat, wtloss.cs.cat)
```




\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{CS vs. UN} \\
    \hline \hline
     & -2 REML & Number of \\ 
    Structure & Log-Likelihood & Cov. Parameters \\ \hline
    Compound Symmetry & `r -2*(-952.9116)` & 10 \\
    Unstructured & `r -2*(-950.4383)` &  18 \\
    Difference & `r -2*(-952.9116) -(-2*(-950.4383)) ` & 8 \\
    \hline
  \end{tabular}
\end{center}

LRT yields $G^2$ =  `r -2*(-952.9116) -(-2*(-950.4383)) ` with 8 df (p =0.7633), so we do not reject the null hypothesis
at $\alpha = 0.05$ and conclude that the assumption of compound symmetry
covariance structure is adequate for the data.



### Unstructured vs autoregressive
$H_0:$ autoregressive is adequate for the data \newline
$H_1:$ unstructred is required

```{r}
anova(wtloss.un.cat, wtloss.ar1.cat)
```

\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{AR-1 vs. UN} \\
    \hline \hline
     & -2 REML & Number of \\ 
    Structure & Log-Likelihood & Cov. Parameters \\ \hline
    Autoregressive & `r -2*(-965.6234)` & 10 \\
    Unstructured & `r -2*(-950.4383)` &  18 \\
    Difference & `r -2*(-965.6234) -(-2*(-950.4383)) ` & 8 \\
    \hline
  \end{tabular}
\end{center}

LRT yields $G^2$ =  `r -2*(-965.6234) -(-2*(-950.4383)) ` with 8 df (p =0.0002), so we  reject the null hypothesis at $\alpha = 0.05$ and conclude  that the assumption of autoregressive covariance structure is \textcolor{red}{inappropriate, worse?}  when compared to unstructured


###Autoregressive vs compound symmetry
```{r}
anova(wtloss.cs.cat, wtloss.ar1.cat)
#same number of parameters so we compare -2*log(likelihood)
```

Since CS and AR-1 have the same number of parameters = 10 , no LRT is necessary.  We can directly compare their likelihoods, or -2*log(likelihood):
\begin{itemize}
\item $-2*log(likelihood)$ for CS = `r -2*(-952.9116)`
\item $-2*log(likelihood)$ for AR-1 = `r -2*(-965.6234) `
\end{itemize}
Since $-2*log(likelihood)$ for CS is smaller  than for AR-1, CS has a higher likelihood
and we conclude that CS is an  adequate model for the covariance
structure when compared to AR-1.

\textbf{The most adequate covariance structure is compound symmetry model.}

\textcolor{red}{Are all these tests correct?} \newline
anova(wtloss.un.cat, wtloss.ar1.cat) \newline
anova(wtloss.un.cat, wtloss.cs.cat) \newline
anova(wtloss.ar1.cat, wtloss.cs.cat) \newline




\begin{center}
  \begin{tabular}{ l  c c c }
    \hline \hline
     & -2 REML & Number of  &\\ 
    Structure & Log-Likelihood & Cov. Parameters  & AIC\\ \hline
    Compoud Symmetry &`r -2*(-952.9116)`& 10 & 1925.823 \\
    Autoregressive & `r -2*(-965.6234)` & 10 & 1951.247\\
    Unstructured & `r -2*(-950.4383)` &  18 & 1936.877 \\

    \hline
  \end{tabular}
\end{center}

Thus, we will use a compound symmetry covariance structure for the remainder of the lab

\newpage

### AUC - test for equality of the area under the curve in two groups.

\newpage

### Parametric curves 
#### Quadratic time trend

```{r}
## QUADRATIC time, CS, REML,
wtloss.cs.quad <- gls(wgtloss~month*prog.fac + I(month^2)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.quad)

## CATEGORICAL time, CS, ML
wtloss.cs.cat.ml <- gls(wgtloss~factor(month)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.cat.ml)
```

$H_0:$ quadratic model \newline
$H_1:$  saturated mode

```{r}
anova(wtloss.cs.cat.ml, wtloss.cs.quad)
```

\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{Testing the Quadratic trend} \\
    \hline \hline
     & -2 Log & Number of \\ 
    Structure & Likelihood & Cov. Parameters \\ \hline
    Quadratic model & `r -2*(-967.85)` & 8 \\
    Saturated model & `r -2*(-967.6657)` &  10 \\
    Difference & `r -2*(-967.85) -(-2*(-967.6657)) ` & 2 \\
    \hline
  \end{tabular}
\end{center}

LRT yields G2 =0.369 (p = 0.8316), so we fail to reject the null hypothesis at
$\alpha = 0.05$ and conclude that the model with Month as a quadratic effect
seems to fit the data adequately. (Note: $chi_{2,0.95}^2 = 5.99$)

#### Linear time trend

```{r}
wtloss.cs.lin <- gls(wgtloss~month*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.lin)
```



$H_0:$ linear model \newline
$H_1:$ quadratic model



```{r}
anova(wtloss.cs.lin, wtloss.cs.quad)
```

\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{Linear vs. Quadratic} \\
    \hline \hline
     & -2 Log & Number of \\ 
    Structure & Likelihood & Cov. Parameters \\ \hline
    Linear model & `r -2*(-968.1477)` & 6 \\
    Quadratic model & `r -2*(-967.85)` & 8 \\
    Difference & `r -2*(-968.1477) -(-2*(-967.85)) ` & 2 \\
    \hline
  \end{tabular}
\end{center}

LRT yields G2 =0.596 (p = 0.7425), so we fail to reject the null hypothesis at
$\alpha = 0.05$ and conclude that the model with Month as a linear effect
fits the data better than the model with Month as a quadratic effect. (Note: $chi_{2,0.95}^2 = 5.99$)






#### Testing for intersections 

```{r}
## Linear model, NO interacations
wtloss.cs.lin.noint <- gls(wgtloss~month + prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.lin.noint)
```

$H_0: \beta_4 = 0$ \newline
$H_1 \beta_4 \neq 0$

```{r}
anova(wtloss.cs.lin)
```

Since the p-value for month*program is $<0.0001$, we reject the null hypothesis and
conclude that there is an interaction between month and program. \newline

Thus, our final model is the linear model with interaction. \newline

\textcolor{red}{What can you conclude about the two weight programs in terms of their
effectiveness?}

\newpage

#### Interpreting quadratic trends:
$\hat{\beta_1}$ =?? expected mean response at baseline of subjects in program 1. \newline
$\hat{\beta_1}$=?? change in expected response at baseline of subjects in program 2 vs.
program 1.\newline
Rate of change in program 1 is $\beta_2 + 2\beta_3 time_{ij}$ . \newline
Plugging in the above estimates,

\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{Program 1} \\
    \hline \hline
     & Rate of & Expected \\ 
    time & Change & Response \\ \hline
    0 &  &  \\
    1 &  &  \\
    2 &  &  \\
    3 &  &  \\
    \hline
  \end{tabular}
\end{center}


\textcolor{red}{Thus, the mean response for Program 1 ?? (increases/decreases) over time.
What about Program 2?}

