---
title: "Complex Data lab 4"
author: "Anna Zaleska"
date: "29 maja 2018"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ALA)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(nlme)
library(lattice)
#knitr::opts_chunk$set(results = 'asis');
options(xtable.comment=FALSE)
library(xtable)
```




###Spaghetti plot of the heart rate over time in 2 groups
```{r}
hrunbalanced <- read.csv(file = "../data/hrunbalanced.csv", header = TRUE)
hrunbalanced$id <- as.factor(hrunbalanced$id)

library(ggplot2)
p <- ggplot(hrunbalanced, aes(x=time, y=hr, color = id)) + geom_line() 
p <- p + facet_grid(.~drug, shrink = TRUE, as.table = FALSE)
p + stat_summary(aes(group=drug), fun.y=mean, geom="line", colour="black", size = 1)
```


```{r}
## Random intercept
fit.RandInt <- lme(hr ~ time*factor(drug),
random=~1|id,
data=hrunbalanced)
##Random intercept and slope
fit.RandSlp <- lme(hr ~ time*factor(drug),
random=~1+time|id,
data=hrunbalanced)

anova(fit.RandInt, fit.RandSlp)
```

```{r}
varCorr <- VarCorr(fit.RandSlp)
covar <- getVarCov(fit.RandSlp)
```



## Task 1

```{r, results= "asis"}
print(xtable(hrunbalanced[hrunbalanced$id %in% c(1,2,13),], caption = "Observations taken into account"),  sanitize.text.function=function(x){x}, include.rownames=FALSE)
```


```{r}
coefs <- fit.RandSlp$coefficients$fixed
#intercept, drug, time, time:drug
coefs <- matrix(c(coefs[1], coefs[3], coefs[2], coefs[4]), ncol = 1)

ones <- matrix(rep(1,6), nrow=6)
zeros <- matrix(rep(0,6), nrow = 6)
tN <- matrix(c(0,1,5,15,30,60), nrow = 6)
t1 <- matrix(hrunbalanced$time[1:6], nrow = 6)
t2 <- matrix(hrunbalanced$time[7:12], nrow = 6)
t13 <- matrix(hrunbalanced$time[73:76], nrow = 4)
x1 <- cbind(ones,zeros,t1,zeros)
x2 <- cbind(ones,zeros, t2,zeros)
x13 <- cbind(ones[1:4,],ones[1:4,],t13,t13)
xNA <- cbind(ones, zeros, tN, zeros)
xNB <- cbind(ones, ones, tN, tN) 
ZN <- cbind(ones,tN)
Z1 <- cbind(ones,t1)
Z2 <- cbind(ones, t2)
Z13 <- cbind(ones[1:4], t13)
```


#### Calculated means:
```{r, results= "asis"}
means <- data.frame(drugA = as.numeric(xNA%*%coefs),
                    drugB = as.numeric(xNB%*%coefs),
                    Y1 = as.numeric(x1%*%coefs),
                    Y2 = as.numeric(x2%*%coefs),
                    Y13 = c(as.numeric(x13%*%coefs),NA,NA))
print(xtable(means, caption = "Conditional means vectors for considered observations"),  sanitize.text.function=function(x){x}, include.rownames=FALSE)
```

#### Calculated covariances:
```{r, results= "asis"}
getCov <- function(Z){
  Z%*%covar%*%t(Z) + as.numeric(varCorr[3,1])*diag(1,nrow = dim(Z)[1], ncol = dim(Z)[1])
}

covZN <- as.data.frame(getCov(ZN))
names(covZN) <- NULL
print(xtable(covZN, caption = " $Cov(Y_i)$ "),  sanitize.text.function=function(x){x}, include.rownames=FALSE)

covZ1 <- as.data.frame(getCov(Z1))
names(covZ1) <- NULL
print(xtable(covZ1, caption = " $Cov(Y_1)$ "),  sanitize.text.function=function(x){x}, include.rownames=FALSE)

covZ2 <- as.data.frame(getCov(Z2))
names(covZ2) <- NULL
print(xtable(covZ2, caption = " $Cov(Y_2)$ "),  sanitize.text.function=function(x){x}, include.rownames=FALSE)

covZ13 <- as.data.frame(getCov(Z13))
names(covZ13) <- NULL
print(xtable(covZ13, caption = " $Cov(Y_13)$ "),  sanitize.text.function=function(x){x}, include.rownames=FALSE)

```


## Task 2
```{r}
hrunbalanced <- read.csv(file = "../data/hrunbalanced.csv", header = TRUE)
#step one
coef.hr <- matrix(NA,ncol = 4, nrow = 0)
for (i in unique(hrunbalanced$id)) {
  model <- lm(hr~time, data = hrunbalanced, subset = id == i)
  model.drug <- unique(hrunbalanced[hrunbalanced$id == i, "drug"])
  coef.hr <- rbind(coef.hr, c(i, model.drug, coef(model)))
}
colnames(coef.hr) <- c("id","drug","intercept","slope")
coef.hr <- as.data.frame(coef.hr)

```

```{r}
# step two
hr.int <- lm(intercept~drug, data = coef.hr)
hr.slp <- lm(slope~drug, data=coef.hr)
print(summary(hr.int))
print(summary(hr.slp))

```
We can see from models' summaries that p-values for drug effect in intercept and slope are 0.186 and 0.366 respectively. Thus we conclude that there is no signifficant difference neither in mean intercepts nor mean slopes in different drugs groups.



```{r}
info.int <- round(summary(hr.int)$coefficients[,c(1,4)], 4)
info.slp <- round(summary(hr.slp)$coefficients[,c(1,4)], 4)
info.mixed <- round(summary(fit.RandSlp)$tTable[,c(1,5)], 4)
```


\begin{table}[b]
\centering
\caption{ Estimates of coefficients and corresponding p-values in two-stage model and mixed-effects model}\label{tab}
  \begin{tabular}{c c c}
     & Two-Stage & Mixed Effects \\ \hline
    Intercept & $`r info.int[1,1]` (`r info.int[1,2]`)$ & $`r info.mixed[1,1]` (`r info.mixed[1,2]`)$ \\
    Time & $`r info.slp[1,1]` (`r info.slp[1,2]`)$ & $`r info.mixed[2,1]` (`r info.mixed[2,2]`)$\\
    Drug & $`r info.int[2,1]` (`r info.int[2,2]`)$ & $`r info.mixed[3,1]` (`r info.mixed[3,2]`)$ \\
    Time*Drug & $`r info.slp[2,1]` (`r info.slp[2,2]`)$ & $`r info.mixed[4,1]` (`r info.mixed[4,2]`)$ \\
    \hline
  \end{tabular}
  \end{table}

Looking at Table \ref{tab} we can observe some differences in coefficients estimates between the two models. However, the differences do not change the signifficance of p-values. In both models drug and time*drug effects are not signifficant.