---
title: "Complex data - lab2"
author: "Stanisław Wilczyński"
date: "15 maja 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ALA)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(nlme)
```

Questions:
1. what is \textbf{L} for this $H_0$ ? (p.3)
2. What is \textbf{L} for $H_0$ : the mean blood lead level is the same at
week 4 and week 6? (p.3)
3. what is r for the $H_0$ in (1) above? (p.9)
4. However it should not be used to perform LRTs for nested
models for mean.Why? (p.10)
5. What type of test was used? What is our conclusion for testing
parallelism? (p.15)
6. What are our conclusions for testing main effects? What other type
test could be used? (p.16)



1. $(0,0,1,0)$
2. $(0,0,1,-1)$ -  a contrast
```{r}
lead <- read.table(file = "../data/lead.txt", header = FALSE)
## Give names to variables
names(lead) <- c("id", paste("y", 1:4, sep=""))
lead.uni <- data.frame(id=rep(lead$id, each=4),
  y=as.numeric(t(as.matrix(lead[,2:5]))),
  time=rep(c(0,1,4,6)),
  time.cat=rep(1:4))

```


3. Here is the output for the test. The answer is 3.
```{r}
lead.cat.ml <- gls(y~factor(time),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
method = "ML",
data=lead.uni)

lead.cat.no.ml <- gls(y~1,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
method = "ML",
data=lead.uni)

anova(lead.cat.ml, lead.cat.no.ml)
```
```{r}
lead.cat <- gls(y~factor(time),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
data=lead.uni)

lead.cat.no <- gls(y~1,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
data=lead.uni)

anova(lead.cat, lead.cat.no)
```

4. REML can't be used to compare nested models for the means in likelihood ratio tests. The reason is that REML estimates the random effects by considering linear combinations of the data that remove the fixed effects. If the fixed effects are changed two models are not directly comparable anymore. For example in case of simple linear regression the restricted maximum likelihood estimator is $\hat \sigma^2 = \frac{RSS}{n-p}$, which is clearly dependent on the number of regression coefficients.



5. We used Wald test for model parameters. We conclude that there is no group by time effect because p-value for group by time is quite large - we do not reject null hypothesis. 

6. We conclude that only time is significant - there is no group effect. Score test, likelihood ratio test and Wald test can be used exchangebly (???) https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faqhow-are-the-likelihood-ratio-wald-and-lagrange-multiplier-score-tests-different-andor-similar/


Other tasks

```{r}
moo <- read.table(file = "../data/mooAll.txt", header = TRUE)
colnames(moo) <- c("protein", "week", "cow", "diet")
moo.gls.interaction <- gls(protein~factor(week)*factor(diet),
correlation=corSymm(form= ~1 | cow),
weights=varIdent(form= ~1 | factor(week)),
data=moo)
summary(moo.gls.interaction)
anova(moo.gls.interaction)
```
```{r}
moo.gls.fixed <- gls(protein~factor(week)+factor(diet),
correlation=corSymm(form= ~1 | cow),
weights=varIdent(form= ~1 | factor(week)),
data=moo)
summary(moo.gls.fixed)
anova(moo.gls.fixed)
```
```{r}
moo.multi <- NULL
for(cow in unique(moo$cow)){
  p1 <- moo$protein[which(moo$cow==cow & moo$week==1)]
  p2 <- moo$protein[which(moo$cow==cow & moo$week==2)]
  p3 <- moo$protein[which(moo$cow==cow & moo$week==3)]
  p4 <- moo$protein[which(moo$cow==cow & moo$week==4)]
  diet <- unique(moo$diet[which(moo$cow==cow & moo$week==1)])[1]
  moo.multi <- rbind(moo.multi, c(p1,max(p2,3),p3,p4,diet))
} 
moo.barley.means <- apply(moo.multi[which(moo.multi[,5]==1),], mean, MARGIN = 2, na.rm=TRUE)
moo.mixed.means <-  apply(moo.multi[which(moo.multi[,5]==2),], mean, MARGIN = 2, na.rm=TRUE)
moo.lupins.means <- apply(moo.multi[which(moo.multi[,5]==3),], mean, MARGIN = 2, na.rm=TRUE)
moo.means <- data.frame(rbind(moo.barley.means, moo.mixed.means, moo.lupins.means))
colnames(moo.means) <- c("t1", "t2", "t3", "t4", "diet")
moo.means <- melt(moo.means, id.vars = c("diet"))


ggplot(moo.means, aes(x=variable, y=value, group=diet, color=diet)) + geom_line()

```

