---
title: "Complex data - lab2"
author: "Stanisław Wilczyński"
date: "15 maja 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ALA)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(nlme)
```

Questions:
1. what is \textbf{L} for this $H_0$ ? (p.3)
2. What is \textbf{L} for $H_0$ : the mean blood lead level is the same at
week 4 and week 6? (p.3)
3. what is r for the $H_0$ in (1) above? (p.9)
4. However it should not be used to perform LRTs for nested
models for mean.Why? (p.10)
5. What type of test was used? What is our conclusion for testing
parallelism? (p.15)
6. What are our conclusions for testing main effects? What other type
test could be used? (p.16)



1. $(0,0,1,0)$
2. $(0,0,1,-1)$ -  a contrast
```{r}
lead <- read.table(file = "../data/lead.txt", header = FALSE)
## Give names to variables
names(lead) <- c("id", paste("y", 1:4, sep=""))
lead.uni <- data.frame(id=rep(lead$id, each=4),
  y=as.numeric(t(as.matrix(lead[,2:5]))),
  time=rep(c(0,1,4,6)),
  time.cat=rep(1:4))

```


3. Here is the output for the test. The answer is 3.
```{r}
lead.cat.ml <- gls(y~factor(time),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
method = "ML",
data=lead.uni)

lead.cat.no.ml <- gls(y~1,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | factor(time)),
method = "ML",
data=lead.uni)

anova(lead.cat.ml, lead.cat.no.ml)
```
4. REML estimators use different observation(??) 



5,6 - check lecture 5,6

```{r}
moo <- read.table(file = "../data/mooAll.txt", header = TRUE)
print(dim(moo))
colnames(moo) <- c("protein", "time", "cow", "diet")
moo.gls <- gls(protein~factor(time),
correlation=corSymm(form= ~1 | cow),
weights=varIdent(form= ~1 | factor(time)),
data=moo)
summary(moo.gls)
```

