---
title: "Complex Data - lab3"
author: "Stanisław Wilczyński"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ALA)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(nlme)
```

```{r}
wtloss <- read.table("../data/weightloss.dat",
header=F)
## Give names to variables
names(wtloss) <- c("id", paste("y", 1:4, sep=""), "program")
## Univariate format
wtloss.uni <- data.frame(id=rep(wtloss$id, each=4),
wgtloss=as.numeric(t(as.matrix(wtloss[,2:5]))),
program=rep(wtloss$program, each=4),
month=seq(0,9,3),
time.cat=rep(1:4))
wtloss.uni$prog.fac <- factor(wtloss.uni$program, labels=c("1:encourage", "2:none"))
attach(wtloss.uni)
wgt.mean <- tapply(wgtloss, list(month, program), mean)
wgt.sd <- tapply(wgtloss, list(month, program), sd)
wgt.mean
wgt.sd
detach(wtloss.uni)
```
```{r}

library(lattice) # Use lattice package
xyplot(wgtloss~month|prog.fac, type='l',groups=id,data=wtloss.uni)
```


```{r}
library(nlme)

## CATEGORICAL time, UN, REML
wtloss.un.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | month),
data=wtloss.uni)
#summary(wtloss.un.cat)

## CATEGORICAL time, CS, REML
wtloss.cs.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni)
#summary(wtloss.cs.cat)

## CATEGORICAL time, AR1, REML
wtloss.ar1.cat <- gls(wgtloss~factor(month)*prog.fac,
correlation=corAR1(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni)
#summary(wtloss.ar1.cat)
```



\begin{center}
  \begin{tabular}{ l  c c }
  \multicolumn{3}{c}{CS vs. UN} \\
    \hline \hline
     & -2 REML & Number of \\ 
    Structure & Log-Likelihood & Cov. Parameters \\ \hline
    Compound Symmetry & `r -2*(-952.9116)` & 10 \\
    Unstructured & `r -2*(-950.4383)` &  18 \\
    Difference & `r -2*(-952.9116) -(-2*(-950.4383)) ` & 8 \\
    \hline
  \end{tabular}
\end{center}


Unstructured vs compound symmetry
```{r}
anova(wtloss.un.cat, wtloss.cs.cat)
```












Unstructured vs auoregressive
```{r}
anova(wtloss.un.cat, wtloss.ar1.cat)
```

Autoregressive vs compound symmetry
```{r}
anova(wtloss.cs.cat, wtloss.ar1.cat)
#same number of parameters so we compare -2*log(likelihood)
```

AUC - test for equality of the area under the curve in two groups.



Parametric curves - quadratic time trend

```{r}
## QUADRATIC time, CS, REML,
wtloss.cs.quad <- gls(wgtloss~month*prog.fac + I(month^2)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.quad)

## CATEGORICAL time, CS, ML
wtloss.cs.cat.ml <- gls(wgtloss~factor(month)*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.cat.ml)
```


Linear time trend

```{r}
wtloss.cs.lin <- gls(wgtloss~month*prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.lin)
```



Testing for intersections 

```{r}
## Linear model, NO interacations
wtloss.cs.lin.noint <- gls(wgtloss~month + prog.fac,
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=wtloss.uni,
method="ML")
#summary(wtloss.cs.lin.noint)
```


